---
title: "Reinforced random walk"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(reticulate)
# py_install("matplotlib")
```



In this worksheet, we will consider a natural random walk process, that is not Markovian.

The reinforced random walk on the integers is defined as follows:  if you are $i \in \mathbb{Z}$, then you can only go to $i+1$ or $i-1$, and you move to $i+1$ with probability $v(i+1)/[v(i-1) + v(i+1)]$, where $v(j)$ is the total number of times an integer $j$ has already been visited.  Initially, we take $v=1$ everywhere. 

It is a theorem of [Tarres (2004)](https://projecteuclid.org/journals/annals-of-probability/volume-32/issue-3B/Vertex-reinforced-random-walk-on-%e2%84%a4-eventually-gets-stuck-on/10.1214/009117907000000694.full) that the walk will *eventually* get stuck on *five* vertices.  In this worksheet we will try to write some code to illustrate this fact.


# Solutions

We will store the number of visits to positive integers in the vectors *vpos*, store the visits to negative integers in *vneg*, and the visits to zero as *vzero*.  We will start the walk at the origin.

```{r}

step <-function(i,l,r){
  nextstep = i-1
  p = r/(l+r)
  if (rbinom(1,1,p) == 1){nextstep <- i+1}
  nextstep
}

steps<- function(N){
  vpos =NULL
  vneg =NULL
  vzero =1
  path = 0
  wr = 1
  wl =1
  k=0
  while(k <N){
    current= path[length(path)]
    if (current+1 >0 && length(vpos) >= current+1){
      wr = vpos[current+1]
    }
    if (current+1 ==0){
      wr = vzero
    }
    if (current+1 <0){
      wr = vneg[abs(current+1)]
    }
  
     if (current-1 <0 && length(vneg) >= abs(current-1)){
      wl = vneg[abs(current-1)]
    }
    if (current-1 ==0){
      wl = vzero
    } 
    if (current-1 >0){
      wl = vpos[current-1]
    } 
    
    
nextspot = step(current, wl,wr)
an = abs(nextspot)
  path <- c(path, nextspot)
  if (nextspot==0){vzero<- vzero+1}
  if (nextspot <0 && an <= length(vneg)){vneg[an] <- vneg[an] +1  }
  if (nextspot <0 && an > length(vneg)){vneg <-c(vneg, 2)  }
  if (nextspot >0 && an <= length(vpos)){vpos[an] <- vpos[an] +1  }
  if (nextspot >0 && an > length(vpos)){vpos <-c(vpos, 2)  }
  wl=1
  wr=1
  k <-k+1
  }  
list(path, vneg, vzero, vpos)
}
steps(2000)
```


# Endnotes

* Version: `r format(Sys.time(), '%d %B %Y')`
* [Rmd Source](https://tsoo-math.github.io/ucl3/week10-2022-inclass-sols.Rmd)

