---
title: "Data from a MC"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
# py_install("matplotlib")
# py_install("pandas")
```



In this worksheet, we will consider the output of a Markov chain that is stored on a textfile.  We will import this data, and do basics statistics to estimate the stationary distribution of corresponding Markov chain.




*  Suppose you are given the output of a $100000$ steps of a irreducible and aperiodic finite state Markov chain.    Carefully explain how you could estimate the stationary distribution for this Markov chain, and why you estimator is reasonable.  

*  Import the data from the file [markovchain.txt](https://tsoo-math.github.io/ucl3/markovchain.txt)  and use this data and your method above to estimate the stationary distribution.

*  Any ideas for the transition matrix?


# Solutions


*  Since the chain is on a finite state space and irreducible and aperiodic, it has a unique stationary distribution $\pi$, and  we know that a version of the law of large numbers gives that for every state $s$, we have

$$\frac{1}{n} \sum_{i=1}^n \mathbf{1}[X_i = s] \to \pi(s).$$
Thus for each state, we simply need to count the number of occurrences and divide by $100000$, to get its approximate probability under $\pi$.  

*  A quick scan of the file shows there are $4$ states: $1,2,3,4$.  With R, we have:

```{r}
z =read.table("markovchain.txt", sep=",")
z= z[,]

b1 = seq(-1,5, by=1)
hist(z, prob=TRUE, breaks=b1)

sum(z==1)/100000
sum(z==2)/100000
sum(z==3)/100000
sum(z==4)/100000
```

Next we show how this code might be done in Python; you may need to install the package pandas; if you are running this inside R,  you may need to put *py_install("pandas")* where you load the *library(reticulate)*.

```{python}
import pandas as pd
import numpy as np

mydata = pd.read_csv("https://tsoo-math.github.io/ucl3/markovchain.txt", sep =",")

print(mydata)

y = mydata.loc[:, 'x']
print(y)
print(np.mean(y==1))
print(np.mean(y==2))
print(np.mean(y==3))
print(np.mean(y==4))


```

# Endnotes

* Version: `r format(Sys.time(), '%d %B %Y')`
* [Rmd Source](https://tsoo-math.github.io/ucl3/week9-2022-inclass-sols.Rmd)

