---
title: "Basic confidence intervals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

We will derive and simulate some  confidence intervals that are typically encountered undergraduate statistics

## Normal random variables, unknown mean, known variance

### Derivation

Let $X_1, \ldots, X_n$ be independent normal random variables all with mean $\mu$ and variance $\sigma^2$.  Here $\mu$ is unknown, but $\sigma$ is *known*.     Let $\bar{X}$ be the usual sample mean.  We know that 

$$Z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \sim N(0,1).$$ 

Recall that for $\alpha \in (0,1)$, $z_{\alpha}$ is the number such that $\mathbb{P}(Z \geq z_{\alpha}) = \alpha$, so that  $\mathbb{P}( -z_{\alpha/2} \leq Z \leq z_{\alpha/2}) = 1- \alpha$.  

Consider the following calculation


$$
\begin{eqnarray*}
1- \alpha &=& \mathbb{P}( -z_{\alpha/2} \leq Z \leq z_{\alpha/2}) \\
&=&  \mathbb{P}\Big( -z_{\alpha/2} \leq \frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \leq z_{\alpha/2}\Big)  \\
&=& \mathbb{P}\Big( -z_{\alpha/2}\frac{ \sigma} { \sqrt{n}} \leq \bar{X} - \mu \leq z_{\alpha/2}\frac{ \sigma} { \sqrt{n}}  \Big)  \\
&=&  \mathbb{P}\Big( \mu \in \big[\bar{X} - z_{\alpha/2} \frac{\sigma}{\sqrt{n}}, \bar{X} + z_{\alpha/2} \frac{\sigma}{\sqrt{n}} \big] \Big);
\end{eqnarray*}
$$

that is, the probability that the true mean $\mu$ lies in the *random* interval given by 

$$\Big[\bar{X} - z_{\alpha/2} \frac{\sigma}{\sqrt{n}}, \  \bar{X} + z_{\alpha/2} \frac{\sigma}{\sqrt{n}} \Big]$$
is $1- \alpha$.  

This calculation motivates the following definition.   Suppose we actually observe the values of $X_1, \ldots, X_n$, and we find that $\bar{X}= \bar{x}$.  Then we say that the *deterministic* interval given by 

$$\Big( \bar{x} - z_{\alpha/2} \frac{\sigma}{\sqrt{n}}, \ \bar{x} + z_{\alpha/2} \frac{\sigma}{\sqrt{n}} \Big)$$
is a $100(1-\alpha)$\%  (two-sided) *confidence interval* for $\mu$.    Sometimes confidence interval is also expressed compactly as
$$\bar{x} \pm  z_{\alpha/2} \frac{\sigma}{\sqrt{n}}.$$
We remark that once data is observed, and a confidence interval is computed, it is a deterministic interval, and whether or not the true mean or parameter lies in the confidence interval is no longer a probabilistic.   


### Code

Here, we will run repeated experiments to demonstrate that the proportion of times the true mean lies in a $100(1-\alpha)$\% confidence interval is indeed $1-\alpha$.

```{r}
experiment <- function(mu, sigma, n, percent, trials)  # N(mu, sigma); n data points; percent of CI; trials=number of experiments 
{
  alpha = 1- percent/100
  c = qnorm(alpha/2,lower.tail=F) # critical value
  b = c*sigma/sqrt(n) 
  totalinside = 0  # total times the true mean mu lies inside the CI
  for(i in 1:trials){
  inside=0
  x = rnorm(n, mu, sigma) # simulate the data
  xbar = mean(x)
  if(mu < xbar + b && mu > xbar - b   ){inside =1}  # check whether the CI contains mu
  totalinside <- totalinside + inside # update
  }
  totalinside/trials  # output the proportion
}
  
experiment(5,2,12, 92, 20000)

```


## Confidence intervals for the sample variance

Let $X_1, \ldots, X_n$ be independent normal random variables all with mean $\mu$ and variance $\sigma^2$, both of which are *unknown*.    Let $\bar{X}$ be the usual sample mean and $S^2$ be the usual sample variance.  We have the nice fact that
$$C:=\frac{(n-1)S^2}{\sigma^2} \sim \chi^2(n-1).$$

Hence it follows that a $100(1-\alpha)$\%  (two-sided) *confidence interval* for $\sigma^2$ is given by

$$\Big( \frac{(n-1)s^2}{r}, \frac{(n-1)s^2}{\ell}\Big),$$
where $S^2 = s^2$ is an observed sample variance, and we compute two critical values:
$$\mathbb{P}(C >r) = \alpha/2  \text{ and }   \mathbb{P}(C <\ell) = \alpha/2.$$ 

### Code

Here, we will run repeated experiments to demonstrate that the proportion of times the true variance lies in a $100(1-\alpha)$\% confidence interval is indeed $1-\alpha$.


```{r}
experimentchi <- function(mu, sigma, n, percent, trials)  # N(mu, sigma); n data points; percent of CI; trials=number of experiments 
{
  alpha = 1- percent/100
  L = qchisq(alpha/2,n-1) # critical values
  R = qchisq(alpha/2,n-1,lower.tail=F)
  totalinside = 0  # total times the true mean mu lies inside the CI
  sigma2 = sigma^2
  for(i in 1:trials){
  inside=0
  x = rnorm(n, mu, sigma) # simulate the data
  sq = var(x)
  S = sq*(n-1)
  a = S/R
  b = S/L # the confidence interval is (a,b)
  if( sigma2 >a && sigma2 <b   ){inside =1}  # check whether the CI contains sigma^2
  totalinside <- totalinside + inside # update
  }
  totalinside/trials  # output the proportion
}
  
experimentchi(8,3,6, 95, 20000)



```

### Exercises

* In the case of the confidence interval for the true mean of a random sample of normals, consider the case where  $\sigma$ is *unknown*.
* In the case of the confidence interval for the true variance of a random sample of normals, consider the case where $\mu$ is  *known*.


### Endnotes

* Version: `r format(Sys.time(), '%d %B %Y')`
* [Rmd Source](https://tsoo-math.github.io/ucl/CI.Rmd)

<br>
<br>
<br>
