---
title: "In class exercise:  Transition matrices"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2:
    keep_tex: yes
  bookdown::word_document2:
    toc: yes
  pdf_document:
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Gambler's ruin

Recall the Gambler's ruin problem from the last session.   Suppose someone plays the following gambling game.   They bet $1$ pound on a fair flip of a coin and win $1$ pound if it comes up heads, and lose if come up tails.
  They start with $1$ pound, and stop playing once they have reached $5$ pounds or have  no money left.   Express this problem up as a Markov chain.  In particular:
  
  * Specify the state space and transition matrix.
  * Take  high powers of the transition matrix in R; comment on what you see.
  
  
#  Anything is possible
  
Consider the transition matrix $P$ on the states $\{1,2,3,4\}$ given by:  
  
```{r}  
P <- matrix(c(0, 1/2, 1/2, 0, 0, 1/2, 0,0,0,1/2, 1/2,0,0, 1/2, 0, 0,0,0,1/2, 1/2, 1/3, 1/3, 0, 0, 1/3), nrow =5)
P <-t(P)
P
```

* Compute powers of $P$ until you see that all the entries are positive.  

* What does it mean if all the entries are positive?



#  Convex combinations

* Let $\lambda$ and $\mu$ be probability measures on $\Omega$ with events $\mathcal{F}$.     

Recall that $\lambda$ is a probability measure if

   - $\lambda(\Omega) =1$
   - $\lambda(A ) \in [0,1]$ for all events $A \in \mathcal{F}$
   - $\lambda( \bigcup_{i=1} ^{\infty} A_i) = \sum_{i=1}^{\infty} \lambda(A_i)$ for any countable number of pairwise disjoint events $A_1, A_2, \ldots$.

Show that 
$$\rho = (1-t)\lambda + t \mu$$
is a probability measure on $\Omega$ for all $t \in [0,1]$.  

* Let $P$ be a transition matrix on a state space $S$.  Show that if $\lambda$ and $\mu$ are stationary measures of $P$, then any convex combination of $\lambda$ and $\mu$ is a  stationary measure for $P$.   

* Consider the transition matrix on four states given by
$$
\left( \begin{array}{cccc}
1/2 & 1/2 & 0 & 0 \\
1/2 & 1/2 & 0 & 0  \\
0 & 0 & 1/3 & 2/3 \\
0 & 0 & 1/2 & 1/2 \\
 \end{array} \right)
 $$
* Find three stationary measures for $P$.  
*  Does $P$ have finitely many stationary measures?


#  Periodicity

Consider the transition matrix $P$ on the states $\{1,2,3\}$ given by:  
```{r}
P <- matrix( c( 0,0,1,0, 0,0,0,1, 1/2,1/2,0,0, 1,0,0,0), nrow =4)
P <- t(P)
P
```
*  Take powers of $P$, what do you observe?
*  Does $P$ have any stationary distributions?  Explain.
*  Consider the transition matrix $Q = P^2$.  Does $Q$ have any stationary distributions.  Explain.  




#   Version: `r format(Sys.time(), '%d %B %Y')`
  * [Rmd Source](https://tsoo-math.github.io/ucl/class3.Rmd)
<br>
<br>
<br>
