---
title: "Gambler's ruin"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2:
    keep_tex: yes
  bookdown::word_document2:
    toc: yes
  pdf_document:
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{exercise, name ="Simple gambler's ruin"}
Suppose someone plays the following gambling game.   They bet $1$ pound on a fair flip of a coin and win $1$ pound if it comes up heads, and lose if come up tails.
  They start with $1$ pound, and stop playing once they have reached $5$ pounds or have  no money left.    Use R to run simulations to estimate the probability of ending the game with no money left.  
```
<br>


```{solution}
First, we define a function that gives the result of one iteration when you have $x$ dollars.   Then we run this *game* function in a while-loop until we are bankrupt or get to win a five pound note.   We want to run the *game* function with the starting amount of $1$, many times and then take an average.  We use the replicate function.
```

```{r}
play <- function(x){
x <- 2*(rbinom(1,1,0.5)) -1 +x;
x
}
game <- function(x){
while( x >0 && x <5){
x <- play(x);
}
x
}
y = replicate(1000, game(1))
backrupt=sum(y==0)/1000
win=sum(y==5)/1000


backrupt
win
```

The following version of the gambler's ruin problem was Huygen's formulation of a problem discussed in a correspondence between Pascal and Fermat: 
*Each player starts with 12 points, and a successful roll of the three dice for a player (getting an 11 for the first player or a 14 for the second) adds one to that player's score and subtracts one from the other player's score; the loser of the game is the first to reach zero points. What is the probability of victory for each player?*
See [A note on the history of the gambler ruin problem](http://koreascience.or.kr/article/JAKO201311637859390.pdf).


```{exercise}
Using R, estimate the probability that first player is victorious.   
```
Of course you can compute the probabilities of getting $11$ and $14$, but you can electronically roll dice:
```{r}
sample(6, 10, replace=TRUE)
```
<br>




```{exercise, name="Three players"}
Suppose three players each of which start with $5$ points, play a game where they have equal probability of winning. The winner takes $1$ point from each of  the two *losers*.    They stop playing when any at least one player has no more money left.  On average, how long do they play?
```

<br>
For more information on the general case of $n$ players see  [The asymmetric n-player gambler's ruin problem with equal initial fortunes](https://doi.org/10.1016/j.aam.2003.07.005).
<br>
<br>
<br>


Version:  5 October 2020
  
  * The R Markdown source file is available at [link](https://tsoo-math.github.io/ucl/gambler.Rmd)

<br>
<br>
<br>

