---
title: "Stat9 Project ICA2  Assessment form"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Candidate(s) identifier: 

# Project title: 

<hr>
<hr>


# Written report  

## Content of written report

* Amount of work done
* Extent to which the student has demonstrated understanding
* Quality & accuracy of reasoning, validity of interpretation, relevance of conclusions
* Critical appraisal, discussion of limitations & further work
* Clarity of objectives
* Quality of the literature review and background information
* Quality of data organisation & collection (if applicable)
* Quality of programming / use of software (if applicable)

<br>

*  Mark1 (out of 100)
```{r}
Mark1=100
```


## Presentation of written report

* Layout of the report and care in its presentation
* Structure of the report
* Standard of judgement shown in selection of material
* Clarity of expression, readability and coherence
  * Correctness of grammar and spelling
* Adequacy of diagrams, tables and graphs (if applicable)
* Quality of presentation of mathematical material (if applicable)

<br>

* Mark2 (out of 100)
```{r}
Mark2=100
```



## Weighted mark for written report



```{r}
MW=0.6*Mark1 + 0.4*Mark2
MW
```

# Oral-recorded presentation

## Clarity, Interest and Structure

* Clarity of aims (Was it clear what the project was trying to achieve?)
* Selection of material--did the presentation focus on features of the work required to understand: 
  * How the aims were addressed? 
  * What were the *headline messages*?
* Clarity of argument / logical flow (Was there a coherent narrative?)
* Clarity of conclusions (Were there clear conclusions? Did they relate to the project aims?)
* Accessibility to intended audience (Was the material presented at a level comprehensible to the speaker’s peers? )

<br>

* Mark3 (out of 100) 

```{r}
Mark3 = 100
```


## Presentation Skills

* Audience engagement:  Was the speaking audible and confident? Was the inflexion varied and interesting? 

* Visual aids:  Were the slides clearly readable and uncluttered? Were the key messages easy to see? Were graphics and tables appropriately labelled and presented so as to convey at a glance the underlying message?
* Appropriate length: Did the speaker fill the time available, comfortably and without rushing?

<br>

* Mark4 (out of 100)

```{r}
Mark4 =100
```


## Weighted mark for oral presentation


```{r}
MO=0.5*Mark3 + 0.5*Mark4
MO
```



# Final mark  

```{r}
FM = 0.7*MW + 0.3*MO
MO
```

# Feedback


<br>
<br>

