---
title: Introduction to queues
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## What are queues    


*  *Items* (customers, patients, cars) enter a system (check-out, hospital, repair shop) for service.

*  They may have to *wait* in-line, in the queue, to be serviced.

*   When they finally receive *service*, this also takes time.

*  When they are done, they depart the system, this may also take time.

*  Queuing time and service time, (and departure time) are often considered together.

*   We are interested in quantities such as how many items are in a system at a time $t$, average time in the system, and the stability of the queue.


## Some terminology and common assumptions

*  The *discipline* of a queue refers to how items are serviced:  first in first out (FIFO) being a common assumption.

*  Usually one or more servers.

*  Arrival time to the queue are often assumed to correspond to a  renewal process.

*  Service times are often assumed to be i.i.d.\

*  Technical maths tools that are used in analysing queues include finding hidden Markov chains.


## Kendall notation

*  A/S/c; three parameters denoting the inter-arrival times, service times, and the number of servers.

*  Inter-arrival times and service times are i.i.d. and the queue is by default FIFO.

*  M stands for Markov (exponential times)

*  M/M/1 or $M(\lambda)/M(\mu)/1$ refers to a queue with one server, where inter-arrival and service times are exponential (with not necessary the same rate).  The *M* stands for Markov

*  G/G/1 refers to a general (and not necessarily the same) distribution for arrivals and service.


## Code M/M/1

Hidden in an early exercise, we had considered such a queue, with $\lambda = 1$ and $\mu = 1.5$.  We had code that produced a histogram of the total an item spends in such a system.


```{r}
inter=rexp(1000, 1)
arr=cumsum(inter)
service = rexp(1000,1.5)

output <- arr[1] + service[1]
for (i in 1:999){
if (arr[i+1] < output[i]){output <- c(output, output[i] + service[i+1])}
if (arr[i+1] > output[i]){output <- c(output, arr[i+1] + service[i+1])}
output
}
```

## continued


```{r}
time = output - arr
x = seq(0,max(time)+1, by=0.01)
b = seq(0,max(time)+1, by=0.1)
hist(time, prob=T, breaks=b)
curve(dexp(x,0.5), add=T)
```


## General results

The ratio of the expected service time over expected inter-arrival time is the **traffic intensity**.   Let  $Q(t)$ be the number of customers in the system (waiting and being serviced) at time $t$.   We say that $Q$ is **stable** if the law of $Q(t)$ converges to a probability distribution as $t \to \infty$.   For a G/G/1 queue, we have:

*  If the traffic intensity is strictly less than $1$, then $Q$ is stable

*  If the traffic intensity is greater than $1$, then $Q$ is not stable

*  If the traffic intensity is $1$, then $Q$ is not stable, provided one of the service time or inter-arrival time is not deterministic.  


## Little's law;  $L = \lambda W$;  

Exact version in $[0,T]$:   *For a system that is empty at $0$ and $T$*



*  $Q(t)$ = number of items in the system at time $t$

*   $\lambda$ = average arrival rate in $[0,T]$ 

*  $N$ = the total number of items arriving  in $[0,T]$

*  $L$ = average number of items in the system in $[0,T]$

*  $W$ =average waiting time of an item in $[0,T]$ 

*  $A = \int_0^T Q(t)dt$ area under $Q$ curve


## Proof

Here *average* is defined so that

*  $L = A/T$

*  $\lambda = N/T$

*  $W = A/N$

*  Hence 
$$L = (A/T) = (A/N) * (N/T) = \lambda W.$$


## Notes

*  Little advocates for this version of his law.  

*  Basically no assumptions required.

*  The quantities $L$, $\lambda$, and $W$ are locked-in a relationship like Ohm's law.

*  One quantity can be easily calculated assuming the others can be measured.  


## Long-term average version via  renewal-reward 

*  Customers arrive one at a time with the $n$th customer spending time $V_n$

*  There exists a **regeneration time** $T_1$ that is finite almost surely, so that the queue is empty, and restarts a fresh at this random time, like a stopping time.

*  Thus the inter-arrival times of these regeneration times $X_i = T_i - T_{i-1}$ are i.i.d. and the regeneration times form a renewal process.  Take $T_0 = 0$. 

* Each time *cycle* $[T_{i-1}, T_i)$ behaves independently. That is, if $N_1$ is the number of arrivals in $[T_0, T_1)$ then it is independent of $N_2$, but identically distributed.

*  We assume that $\mathbb{E} N_1 < \infty$, $\mathbb{E} T_1 < \infty$, and $\mathbb{E}(N_1 T_1) < \infty$.


##  Continued

*  Consider the reward at each cycle to be given by

$$ R_i = \int_{T_{i-1}} ^ {T_{i}} Q(t)dt$$

*  Clearly, the $R_i$ are i.i.d. and $\mathbb{E} R_1 \leq \mathbb{E}N_1T_1 < \infty$.

*  By the law of large numbers for renewal processes it follows that if $C(t)$ is the total reward up to time $t$, we have 

$$ \lim_{t \to \infty} \frac{1}{t}\int_0 ^t Q(u)du = \lim_{t \to \infty} C(t)/t  = \mathbb{E} R_1 / \mathbb{E} T_1=:L$$

## continued


*  Similarly, give ourselves the reward $N_i$ on each cycle, the total of number of items $N(t)$ satisfies

$$ N(t)/t \to \mathbb{E} N_1 / \mathbb{E} T_1:= \lambda$$

## continued

*  Consider another inter-arrival process with inter-arrival times $N_i$ and give ourselves the reward of the total waiting time in the $i$th cycle; in particular, $S_1 = \sum_{k=1}^ {N_1} V_k$, and the $S_i$ are i.i.d. with 

$$ \mathbb{E}S_1  \leq \mathbb{E} N_1 T_1 < \infty$$

*  Again, by the renewal-reward theorem, we have

$$\lim_{n \to \infty} \frac{1}{n}\sum_{k=1} ^n V_k \to \mathbb{E} S_1 / \mathbb{E} N=:W$$  

##  Little's theorem $L = \lambda W$.

Observe that 

$$\mathbb{E}  S_1= \mathbb{E} \Big( \sum_{k=1} ^{N_1}  V_k  \Big) = \mathbb{E} \Big( \int_0 ^{T_1} Q(u)du \Big) = \mathbb{E} R_1$$
Thus
$$
\begin{eqnarray*}
 L &=& ( \mathbb{E} R_1 /\mathbb{E} T_1 ) \\
 &=&  ( \mathbb{E} R_1 /\mathbb{E} N_1 )*  (\mathbb{E} N_1 / \mathbb{E} T_1) \\
 &=&  (\mathbb{E} S_1  / \mathbb{E} N_1) * \lambda \\
 &=&  \lambda W.
 \end{eqnarray*}
$$

## Summary

* We introduced queues and [Kendall's notation](https://projecteuclid.org/euclid.aoms/1177728975)

*  We discussed [Little's theorem](https://pubsonline.informs.org/doi/abs/10.1287/opre.1110.0940).  The link requires UCL VPN.


##  Version: `r format(Sys.time(), '%d %B %Y')`
  * [Rmd Source](https://tsoo-math.github.io/ucl/intro-queues.Rmd)



<br>
<br>
<br>


