---
title: "Introduction to R"
output: ioslides_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is R?

- R is a free open-source program that is 
- Used to do statistics in both academia and industry.   
- Many online resources for R, for both beginners and experts.
- Good skill to put on your cv.

## A fancy calculator

R can be used as a fancy calculator
```{r}
2+2
sin(3.14)
exp(1)
```
##
```{r}
log(2.71)
pi
```
##
Variables can be assigned in the following way

```{r}
x <- 1+2+3+4+5+6
x*2
```


## Entering data by hand into R

R is designed to store data as vectors $x = (x_1, x_2, \ldots, x_n)$ that is lists of numbers.  

```{r}
y <- c(1,2,3,4,5,6,7,8,9)
y
```

## Basic operations with vectors

R has many built in common operations that are useful for statistics:

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
x
sum(x)
```
##
```{r}
mean(x)
sd(x)
var(x)
```

## 

R will do certain operations component wise:
```{r}
 x <- c(1,2,3,4,5)
y <- c(6,7,8,9,10)
z=x+y
z

sin(z)

```

## 

It is often necessary to add or delete data from a vector:
```{r}
x<- c(0.1, 0.2, 0.3, 0.4, 0.5)
x <- x[-5]
x
x <- c(x, 5.5)
x
```

##  Random variables in R

- R can be used to simulate coin flips and dice roll.   
- R does not use true randomness.  
```{r}
x<- rbinom(12,1, 0.5)
x
z <- sample(6,12, replace =TRUE)
z
```
##
```{r}
x<- rnorm(10, 5, 1)
x

z <- runif(10, min=-1, max=1)
z
```
##
```{r}
z <- rexp(10, 2)
z
z<- rpois(7, 1)
print(z)
```  
## Histograms in R

```{r}
x <- rexp(100,1)
hist(x, prob=TRUE)
```


## Functions

- We might need to compute  $\sin(x) + \cos(x)$ more than once. 
```{r}
sincos <- function(x){
z <- sin(x) + cos(x);
z
}
sincos(1)
```
<br>


- It is sometimes useful have functions that do not take inputs
- Roll a fair dice 10 times, and take the sum





## Basic programming in R

- We will introduce basic programming in R with the following problem
- We will use a *while* loop in R.


By running simulations in R, approximate the average number of  rolls of a fair dice it takes before you see a 6.

## The procedure


- We make a function *numrolls* which counts how many times we need to roll a dice until we see a $6$  
- Use the *replicate* command to repeat this function many times
- Take the average

## The code


```{r}
numrolls <- function(){
n=0
x=0
while(x <6){
x <- sample(6,1, replace =TRUE)
n <- n+1
}
n
}

mean(replicate(1000, numrolls()) )
```

##  For loop

We illustrate how to write a *for* loop in the context of prime numbers.


Define a function that tells you whether a positive integer is prime or not.

## Background

- Let $n$ be an integer.  

- $d$ is a divisor of $n$ if  there exists an integer $c$ such that $n = cd$.  

- An  integer $n \geq 2$ is prime if it only divisors are $1$ and $n$.   

- R has a built in remainder function, which for nonnegative integers $a,b$ outputs the remainder in the sense of elementary school, when $a$ is divided by $b$.  

## Procedure


Using the remainder function we define the *isprime* function, and use it spit out the prime numbers up to 500.

```{r}
25%%5
26%%5
```

## Code

```{r}
isprime <- function(n){
x=1
for (i in 2: (n-1)){
if (x >0) {
x <- n%%i
}
}
if (n==1) {x <-0}
if (n==2) {x <-1}
 x
 }
isprime(101)
```
##
```{r, R.options = list(width = 50)}
x=2
 for(i in 3:500){
if( isprime(i)==1){
x <- c(x, i)}
 x
}
print(x)
```

## Summary

- We introduced some basics of R
- We gave examples programming basics 
- Functions
- While loops 
- for loops 
- the replicate function.



## Version:  4 October 2020



<br>
<br>
<br>


