---
title: "Introduction to R"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2:
    keep_tex: yes
  bookdown::word_document2:
    toc: yes
  pdf_document:
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is R?


R is a free open-source program that is used to do statistics in both academia and industry.   There are many online resources for R, for both beginners and experts.    For example, [Venables, Smith, and the R Core Team](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)

## A fancy calculator

R can be used as a fancy calculator
```{r}
2+2
sin(3.14)
exp(1)
log(2.71)
pi
```
Variables can be assigned in the following way

```{r}
x <- 1+2+3+4+5+6
x*2
```


## Entering data by hand into R

R is designed to store data as vectors $x = (x_1, x_2, \ldots, x_n)$ that is lists of numbers.  

```{r}
y <- c(1,2,3,4,5,6,7,8,9)
y
```

## Basic operations with vectors

R has many built in common operations that are useful for statistics:

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
x
sum(x)
mean(x)
sd(x)
var(x)
```
R will do certain operations component wise:
```{r}
 x <- c(1,2,3,4,5)
y <- c(6,7,8,9,10)
z=x+y
z

x*y

sin(z)
```
It is often necessary to add or delete data from a vector:
```{r}
x<- c(0.1, 0.2, 0.3, 0.4, 0.5)
x <- x[-5]
x
x <- c(x, 5.5)
x
```

##  Coin-flips, dice roll, normal, and uniform random variables in R

R can be used to simulate coin flips and dice roll.   R does not use true randomness, rather it generates coin clips using a deterministic algorithm that simulates true randomness.  
```{r}
x<- rbinom(12,1, 0.5)
x
z <- sample(6,12, replace =TRUE)
z
x<- rnorm(10, 5, 1)
x
z <- runif(10, min=-1, max=1)
z
```  
# Histograms in R

```{r}
x <- rexp(100,1)
hist(x, prob=TRUE)
```
<br>
Here $x$ is a 100 randomly generated data points from the exponential distribution with rate $1$.


# Functions

Suppose we needed to use the quantity $\sin(x) + \cos(x)$ over and over again for different values of $x$, then it may be useful to define this as a function in the following way:
```{r}
sincos <- function(x){
z <- sin(x) + cos(x);
z
}
sincos(1)
```
<br>
Here the function sincos takes an input  $x$.  It is sometimes useful have functions that do not take inputs, but simple perform operations:  say roll a fair dice 1o times, and take the sum.





# Basic programming in R

We will introduce basic programming in R with the following exercise and will illustrate how to write a *while* loop in R.
<br>
```{exercise}
By running simulations in R, approximate the average number of  rolls of a fair dice it takes before you see a 6.
```
<br>
```{solution}
We make a function *numrolls* which counts how many times we need to roll a dice until we see a $6$.  Then we use the *replicate* command to repeat this function many time, and take the average.
```
<br>
```{r}
numrolls <- function(){
n=0
x=0
while(x <6){
x <- sample(6,1, replace =TRUE)
n <- n+1
}
n
}

mean(replicate(1000, numrolls()) )
```


In the next exercise, we illustrate how to write a *for* loop.

```{exercise}
Define a function that tells you whether a positive integer is prime or not.
```
<br>
```{solution}
Let $n$ be an integer.  We first recall that $d$ is a divisor of $n$ if  there exists an integer $c$ such that $ n = cd$.  An  integer $n \geq 2$ is prime if it only divisors are $1$ and $n$.   R has a built in remainder function, which for nonnegative integers $a,b$ outputs the remainder in the sense of elementary school, when $a$ is divided by $b$.  Using the remainder function we define the *isprime* function, and use it spit out the prime numbers up to 500.

```{r}
25%%5
26%%5

isprime <- function(n){
x=1
for (i in 2: (n-1)){
if (x >0) {
x <- n%%i
}
}
if (n==1) {x <-0}
if (n==2) {x <-1}
 x
 }
 isprime(1)
 isprime(2)
isprime(101)

x=2
 for(i in 3:500){
if( isprime(i)==1){
x <- c(x, i)}
 x
}
x
```

# Summary

We introduced some basics of R, and we gave examples programming basics including defining functions, while loops, for loops, and using the replicate function.

<br>
<br>

# Version:  25 September 2020



<br>
<br>
<br>


