---
title: "Homework 4"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2:
    keep_tex: yes
  bookdown::word_document2:
    toc: yes
  pdf_document:
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{exercise, name="Deterministic colouring"}
Let $\Pi$ a Poisson point process on $[0, \infty)$.  Suppose we colour the first arrival blue, and then next arrival red, and continue colouring the points in this alternating fashion.   Consider the point processes $\Gamma$  formed by considering only the blue points.  Is this a Poisson point process?  Explain.
```
<br>
```{solution}
*No*.  The interarrival times between two blue points is given by the sum of *two* exponential random variables, and is no longer an exponential. 
<br>

```{exercise, name="Shop keeper"}
Suppose we model the number of customers that arrive at a high street shop on at particular day by a Poisson process of intensity $\lambda >0$, where $\lambda$ is measured in customers per hour.     We wish to estimate $\lambda$.    Suppose the shop is really high-end and on some days has no customers, on its $6$ hours of operations.  The shop keeper only keeps track of whether she had has any customers are not; that is, her records $x = (x_1, \ldots, x_n)$ are a binary sequence.  Find a consisent estimtor for $\lambda$

```
<br>
```{solution}
Let $X_i$ be the binary Bernoulli random variables that corresponds to whether a customer arrived on the $i$th day.  Note that $\mathbb{P}(X_i = 0) = e^{-6\lambda}$ and $\mathbb{P}(X_i = 1) = 1 - e^{-6\lambda}$.  Assume that these are independent random variables, the law of large numbers gives that
$$ n^{-1}(X_1 + \cdots X_n)  \to \mathbb{E}X_1 = 1 - e^{-6\lambda},$$
  as $n \to \infty$.
  Thus setting   $$T_n :=  -\frac{1}{6}\log( 1- n^{-1}(X_1 + \cdots+ X_n)),$$
  we also know that $T_n \to \lambda$, as $n \to \infty$, since $\log$ is continuous.    
  <hr>
  ```{exercise, Name="Uniforms"}
Suppose we $\Pi$ is a Poisson point process on $[0, \infty)$ of intensity $\lambda$.     Using the construction of $\Pi$ as exponential inter-arrival times, prove that conditioned on the event that the unit interval contains exactly one point, the distribution of the its location is uniform.    
  ```
  <br>
```{solution}
Let $X_1$ and $X_2$ be independent exponential random variables of rate $\lambda$ representing the inter-arrival times.  Let $U$ be the location of the point in question.  We are asked to compute
$$\mathbb{P}(U \leq x | X_1 < 1, X_2+X_1 >1)=\mathbb{P}(X_1 \leq x | X_1 < 1, X_2+X_1 >1)$$
  for $0 <  x <1$.  
Since $X_1$ and $X_2$ are independent, we have
\begin{eqnarray*}
\mathbb{P}(X_1 < x, X_2+X_1 >1) &=&  \int_0 ^x  \lambda e^{- \lambda x_1} \Big( \int_{1-x_1} ^{\infty}   \lambda e^{ -\lambda x_2} dx_2 \Big)dx_1  \\
 &=&\int_0 ^x \lambda e^{-\lambda x_1} e^{-\lambda(1-x_1)}dx_1  \\
&=& x \lambda  e^{-\lambda}.
\end{eqnarray*}
Hence, using this formula in the numerator, and also in the denominator with $x=1$, we have
$$\mathbb{P}(U \leq x | X_1 < 1, X_2+X_1 >1) =  \frac{x \lambda  e^{-\lambda}}{  \lambda  e^{-\lambda}} =x$$
  as desired.
```




```{exercise, name="Random deletion"}
Simulate a Poisson process of intensity $\lambda=2$, say with $10000$ arrivals.  Delete each arrival independently with probability $p=\tfrac{1}{2}$ to from a new *thinned* process.  Plot a histogram of  the inter-arrival times of the thinned process.  What should you see?  Why?
```


```{solution}
The histogram approximates the exponential density with rate $1$, since we know that the resulting thinned process is still a Poisson process with intensity $p \lambda=1$.
```

```{r}
inter=rexp(10000, 2)
arr = cumsum(inter)
coin = rbinom(length(arr), 1, 0.5)
for (i in 1:length(arr)) if (coin[i]==0){arr[i]<- 0}
arr<-setdiff(arr, 0)
interthin = arr[1]
for (i in 1: length(arr)-1){interthin <- c(interthin, arr[i+1] - arr[i]) }
x = seq(0, max(interthin)+1, by=0.1)
hist(interthin, prob=T, x)
curve(dexp(x), add=T)

```

```{exercise, name="Poisson on a disc"}
Simulate a Poisson point process of intensity $100$ on a disc.
```

```{solution}
We simulate a single point on the disc, by simulating a uniform point on a square, and keep it if lands inside an inscribed disc, or repeat otherwise.  Then we implent the characterization of Poisson point processes as a Poisson number of uniform points.   Note a disc has area $\pi r^2$.

```

```{r}
point <- function(){
  z=2
  while(length(z)==1){
  x = 2*runif(1) -1
  y = 2*runif(1) -1
if (x^2 + y^2 <1) { z<-c(x,y)}
  }
  z
}
 re=replicate(rpois(1,100*pi), point())
 plot(re[1,], re[2,], xlim=c(-1.1,1.1), ylim=c(-1.1,1.1), asp=1)
```


