---
title: "UCL STATISTICAL SCIENCE POSTER DAY"
output:
  html_document: default
  pdf_document: default
---

\pagenumbering{gobble}


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r, echo=FALSE, results= 'hide', warning = FALSE}
library(latex2exp)
```


```{r, echo=FALSE}
x1 = c(10,8,13,9,11,14,6,4,12,7,5)
y1 = c(8.04,6.95,7.58,8.81,8.33,9.96,7.24,4.26,10.84,4.82,5.68)

x2 = c(10,8,13,9,11,14,6,4,12,7,5)
y2 = c(9.14,8.14,8.74,8.77,9.26,8.10,6.13,3.1,9.13,7.26,4.74)

x3 = c(10,8,13,9,11,14,6,4,12,7,5)
y3 = c(7.46,6.77,12.74,7.11,7.81,8.84,6.08,5.39,8.15,6.42,5.73)

x4 = c(8,8,8,8,8,8,8,19,8,8,8)
y4= c(6.58,5.76,7.71,8.84,8.47,7.04,5.25,12.5,5.56,7.91,6.89)



# bmp(filename="ans.bmp", width=960)
par(mfrow=c(2,2), mar = c(5, 5, 5, 5), pch=19, cex=0.7)

plot(x1,y1, col = "blue",  pch=20, xlim=c(4,19), xlab = "Prizes", ylab="All PhD students welcome", axes=FALSE, main="27 OCTOBER", cex.main = 2, asp=1)





plot(x2,y2, col = "red",  pch=20, xlim=c(4,19), xlab = TeX(r'($\bar{x} = 9$)'), ylab=TeX(r'($\bar{y} = 7.50$)'), axes=FALSE, main="Visit & Present Your Work", cex.main = 2,asp=1)



plot(x3,y3, col = "dark green",  pch=20, xlim=c(4,19), xlab = TeX(r'($s_{x}^2 = 11$)'), ylab=TeX(r'($s_y^2 = 4.125$)'), axes=FALSE, main="",  cex.main = 2,asp=1)



plot(x4,y4, col = "purple",  pch=20, xlim=c(4,19), xlab = "Please contact Terry Soo; t.soo@ucl.ac.uk", ylab=TeX(r'($\rho =0.816$)'), axes=FALSE, main="",  cex.main = 2,asp=1)
curve(3 + 0.5*x, add=TRUE)
```

### Additional notes and updates


*   3.30 - 5.30 PM, at [1-19 Torrington Place](https://www.ucl.ac.uk/maps/1-19-torrington-place), 1st Floor,  UCL Stats Department

*   Guest Judge Prof [Jim Smith](https://warwick.ac.uk/fac/sci/statistics/staff/academic-research/smith/) 
    * Prof Smith will also be giving the departmental seminar at 2 PM
    
* Cash prize:  £ 500 
  * Kevin Huang (UCL Gatsby) and Ilina Yozova (UCL Stats) [shared prize]

* Book prize:  Generously provided by Cambridge University Press
  * David Bang (Warwick), Max Hird (UCL Stats), and Luke Hardcastle (UCL Stats) 

*  Last updated: `r format(Sys.time(), '%d %B %Y')`




```{r, echo=FALSE, results='hide'}
dev.off()
```




