---
title: "Non-homogeneous Poisson processes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

For many applications, it is obvious that one should consider a rate $\lambda = \lambda(t)$ that depends on time; the expected number arrivals during to a coffee shop will most likely depend on the time.  Here we are interested in simulating such Poisson processes by thinning as constructed by [Lewis and Shedler, 1979](https://onlinelibrary.wiley.com/doi/abs/10.1002/nav.3800260304).  Recall that if we are given a Poisson process of intensity $4$, and we *keep* each arrival independently with probability $\tfrac{2}{3}$, we are left with a Poisson process of intensity $\tfrac{8}{3}$.  This same idea can be used to generate a non-homogeneous Poisson point process of intensity $\lambda(t)$, where $0\leq \lambda(t) \leq M$ is uniformly bounded by $M$.  

  *  First generate a homogeneous Poisson process of intensity $M$
  *  At each point/arrival $t$ that is generated, we keep it independently with probability 
  $$ \frac{\lambda(t)}{M}$$

## Exercise 

Suppose that we are given $\lambda(t) = \sin(t) + 4$.  Generate a nonhomogeneous Poisson point process on $[0, 200]$ with intensity function $\lambda$. 


## Solution

First, we generate the homogeneous Poisson point process of intensity $5$.

```{r}
arr <- function(end){
T = rexp(1,5)
while(T[length(T)] < end){
  T <- c(T,  (T[length(T)] + rexp(1,5))  )
}
T <- T[-length(T)]   # remove the last arrival, as it will be larger than 10
T
}

z=arr(10)  # outputs the arrival times of a Poisson process up to time 10
```
Next, we write a function that applies the thinning

```{r}
thinning <- function(x){
  del=NULL
  for( i in 1:length(x) ){
    t= x[i]
    prop = (sin(t)+4)/5
    coin = rbinom(1,1,prop)
    if(coin==0){
      del = c(del, i)
    }
  }
  del    # outputs the coordinates to be deleted.
}
z
thin=thinning(z)
nonhomg = z[-thin]    # the corresponding entries deleted.
thin
nonhomg
length(z)
length(thin)
length(nonhomg)
```

