---
title: "In class exercise:  Return times and symbolic computing"
output:  html_document
---


```{r setup, include=FALSE}
library(reticulate)
# py_install("matplotlib")
# py_install("sympy")
knitr::opts_chunk$set(echo = TRUE)
```


# Symbolic computing inside Python  (inside R)

Will we explore basic symbolic computing inside Python.  In symbolic computing, the software is able to compute $2x = x+x$  or $\int_0^x x^2 dx = \tfrac{x^3}{3}$ *without* having a numerical value for $x$.  Popular software includes, Maple, Mathematica, and Sage; Python's *sympy* is a basic in-house substitute, which we will explore.


```{python}
import sympy as sym

x = sym.pi   # the digit pi
print(x)
print("#############")
print(x.evalf())
print("#############")
print(x.evalf(100))
print("#############")

x = sym.Symbol('x')   # you got to tell Python that x is symbol
print(sym.simplify((x + x)))

y = sym.Symbol('y')
y=sym.integrate(3 * x ** 2, x)  # integrate 3x^2
print(y)
print("#############")
```

We can also do linear algebra

```{python}
a = sym.Symbol('a')
b = sym.Symbol('b')
c = sym.Symbol('c')
d = sym.Symbol('d')
M =  sym.Symbol('M')
M=sym.Matrix([[a, b], [c, d]])
print(M)
print("#############")
print(M.det())
print("#############")
print(M.T)
print("#############")
print(M.eigenvects())
print("#############")
Msub = M.subs(a,1).subs(b,1).subs(c,1).subs(d,0)
values = Msub.eigenvects()
print("#############")
print(values)
print("#############")


```

The output may be a bit hard to read, we have the eigenvalue, its multiplicity, and the eigenvector(s): here we substitute in the identity matrix in two dimensions, so that the eigenvalue $1$ has en eigenbasis that is spanned by the usual coordinate basis vectors.

```{python}
M=sym.Matrix([[a, b], [c, d]])
Msub = M.subs(a,1).subs(b,0).subs(c,0).subs(d,1)
values = Msub.eigenvects()
print(values)
```


# A Markov chain

Consider the three state Markov chain on $\{1,2,3\}$ where 
$$p_{11} = p = 1-p_{12}$$ 
$$p_{22} = q=1- p_{23}$$
and
$$p_{31}=1.$$

We will consider the case where $q \in (0,1)$ is fixed, but will think of $p \to 1$.  

* Write down the transition matrix.

* Suppose the Markov chain $X$ with these transition probabilities starts in state $1$.  Let $T = \inf\{n \geq 1: X_n = 1\}$.   Compute (by brute force) $\mathbb{E}(T)$.

* Find the stationary distribution of the transition matrix.

*  Simulate the chain, and leave $p$ and $q$ as variables that you can choose.

*  See that all your findings and answers are consistent.

## Partial solutions

In order to compute $\mathbb{E}(T)$, we can save a little time, by employing the strong Markov property. Let $G$ be geometric random variable  with parameter $1-q$ that is independent of the chain $X$;  note that $1-q$ is the probability that we can exit state $2$, given we are there.  Then $T$ has the same law as:

$$T' = 1 \cdot \mathbf{1}[X_1=1] + \mathbf{1}[X_1=2](2+G),$$

here recall that $\mathbf{1}[X_1=1]$ is one if $X_1=1$ and zero if $X_1=0$.   
Then by the linearity of expectation and the independence of $G$ and $X$, we have 

$$\mathbb{E}(T') = p + (1-p)(2 + \frac{1}{1-q});$$

notice as $p \to 1$, we have that $\mathbb{E}(T) \to 1$, as expected.

We can compute the eigenvalues of the transition matrix in Python.  We know from theory, that if $\pi$ is the stationary distribution, then $\mathbb{E}(T) = (\pi(1))^{-1}$, which we can confirm here algebraically. 


```{python}
p = sym.Symbol('p')
q = sym.Symbol('q')
P =  sym.Symbol('P')
P=sym.Matrix([[p, 1-p,0], [0,q, 1-q], [1,0,0]])
Pt =  sym.Symbol('Pt')
Pt= P.T
print("#############")
print(Pt.eigenvects())
print("#############")
print(sym.simplify( ( 1/(1-p)  + 1/(1-q) +1  ) / (1/(1-p))         )        )
print("#############")
print(sym.simplify(2-p + (1-p)/(1-q)                )     )
print("#############")
print(sym.simplify(    ( 1/(1-p)  + 1/(1-q) +1  ) / (1/(1-p)) -  (2-p + (1-p)/(1-q)  )        )        )
```




We will simulate the chain in R, for the values $p=0.75$ and $q=0.25$.  We will find the average number of times the chain is in state $1$.  We will also run code to estimate $\mathbb{E}(T)$.

```{r}
step <-function(i,p,q){
P <- matrix(c(p,1-p,0, 
              0,q,1-q, 
              1,0,0), nrow =3)
P <-t(P)
q = P[i,]
  x=-1
  u = runif(1)
  j=0
  cumq = cumsum(q)
  while(x==-1){
    j<-j+1
    if(u <= cumq[j]){x <-j}
  }
  x
}

steps <- function(n,p,q){
  x = 1
  for (i in 1:n){
    x <- c(x, step(x[i],p,q))
  }
  x
}
mc=steps(5000,0.75,0.25)
stat = c(mean(mc==1), mean(mc ==2),mean(mc ==3))
stat

time <- function(i,p,q){
  n =0
  x=i
  n <- n+1
  x <-step(x,p,q)
  while(x !=i){
    n <-n+1
    x <- step(x,p,q)
  }
  n
}

m=mean(replicate(5000, time(1,0.75,0.25)))

1/m


```


```{python}
x=((p + (2 - p)*(q - 1) - 1)/(q - 1)).subs(p,0.75).subs(q,0.25)
y=1/x
print (y.evalf())
```

#  Endnotes

* Version: `r format(Sys.time(), '%d %B %Y')`
* [Rmd Source](https://tsoo-math.github.io/ucl2/returntime-sol.Rmd)
<br>
<br>
<br>
